---
title: "Homework-03"
author: "Callie Baker"
date: "2026-05-28"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

# Set up

## Packages

```{r}
#| label: packages and data

# read in packages
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(ggplot2)
library(MuMIn) # model selection
library(dplyr)

# reading in data
kelp <- read_csv(here("data", "temp-kelp.csv"))
my_data <- read_csv(here("data", "envs193ds-personal-data.csv"))

```

# Problem 1. Giant kelp fronds

## a. 
To determine the strength of the relationship between temperature and giant kelp frond elongation rate, I can run tests for the correlation between the two variables. I can calculate Pearson's correlation (r) or for Spearman rank correlation ($\rho$). Pearson's correlation is used when variables are continuous and normally distributed; Pearson's r represents the linear relationship between two variables and is used for parametric studies. To contrast, Spearman rank correlation is the non parametric version which describes monotonic relationship between variables. 

## b.
```{r}
#| label: visualization

# base layer: kelp data
ggplot(data = kelp,
       # x-axis: temperature
       # y-axis: elongation rate
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # first layer: points to represent individual kelp observations
  # changed point color to dark green 
  geom_point(color = "darkgreen") + 
  # relabeled x- and y-axes
  # added title
  labs(x = "Ocean Temperature (ºC)", 
       y = "Kelp Elongation Rate (cm/day)",
       title = "Kelp Grows Faster at Cooler Ocean Temperatures") +
  # changed theme from default
  theme_minimal()

```

## c. 


## d.


# e. 


## f. 

# Problem 2. Personal data

## a. 

Figure 1. 
```{r}
#| label: my-data-categorical-visualization
# creating boxplot with jitter to visualize workout duration vs workout type

# cleaned column names
my_data <- my_data |> 
  clean_names()

# base layer: ggplot with x- and y-axes
# data frame: my_data
ggplot(data = my_data,
       # x-axis: workout type (categorical predictor variable)
       # y-axis: workout duration (response variable)
       mapping = aes(x = workout_type,
                     y = duration_min,
                     # colored points by workout type
                     color = workout_type)) +
  # first layer: boxplot to show distribution
  geom_boxplot() +
  # second layer: jitter to show underlying data
  geom_jitter(height = 0, # no jitter in vertical direction
              width = 0.2, # narrowed jitter in horizontal direction
              alpha = 0.5, # made points more transparent
              shape = 1) + # changed points to be open circles
  # changed colors from default
  scale_color_manual(values = c("Lift" = "darkgreen",
                                "Run"  = "navyblue",
                                "Yoga" = "darkred")) +
  # relabeled x- and y-axes
  # added title summarizing takeaway
  labs(title = "Runs show the most variable workout duration",
       x = "Workout type",
       y = "Workout duration (minutes)",
       subtitle = "2026-05-26") +
  # changed ggplot theme from default
  theme_minimal() +
  # removed legend
  theme(legend.position = "none")
```

Figure 2. NEED TO UPDATE
```{r}
#| label: my-data-continuous-visualization
# creating scatter plot to visualize workout duration (minutes) vs 
# sleep the previous night (hours)

# base layer: ggplot with x- and y-axes
# data frame: my_data
ggplot(data = my_data,
       # x-axis: sleep (hours) the night prior (continuous predictor
       # variable)
       # y-axis: workout duration (response variable)
       mapping = aes(x = sleep_hours,
                     y = duration_min)) +
  # first layer: points for each observation
  geom_point(size = 2) + # increased point size
  # relabeled x- and y-axes
  # added title summarizing main message
  labs(title = "More sleep may be associated with longer workouts",
       x = "Sleep the previous night (hours)",
       y = "Workout duration (minutes)") +
  # changed ggplot theme from default
  theme_minimal() 
```

## b. 

Captions 

From the boxplot with overlayed jitter (problem 3a), I observed that my run workouts show the greatest variability in duration indicated by the wider spread of points and the taller box. Of the three groups, run workouts have the smallest median duration, while lift workouts have the largest median duration. Yoga workouts shows the most consistency with points clustered closely together at 60 minutes and no vertical box. In the continuous scatter plot showing workout duration vs sleep in hours (problem 3b), there does not appear to be a strong relationship between variables, seen through the varied points. Collecting more data could shift the medians and spreads in the boxplots, and a larger sample size may reveal a clearer relationship between continuous variables in the scatter plot. 

---

